---
title: "Analyzing Workforce Dynamics and Safety: Insights from an Oil Refinery"
date: last-modified
date-format: "[Last modified on] MMMM DD, YYYY HH:mm:ss zzz"
format: 
  html:
    theme: cosmo  # zephyr pulse, sketchy, default, cosmo, vapor etc
author:
  - name: Mariana Solanilla
    email: solanillam@appstate.edu
    affiliation:
      - name: Appalachian State University
        city: Boone
        state: NC  
        url: https://www.appstate.edu/
abstract: >
 This project analyzes two years of data (2022–2023) on employee work hours, overtime, and safety incidents at an oil refinery. Through data cleaning and visualization, key trends and patterns, including workload distribution and incident frequencies, are explored. The findings, presented on an interactive web page, offer insights into workforce dynamics and safety outcomes with recommendations for further consideration.
keywords:
  - R
  - data viz
license: "CC BY"  
code-fold: false
---

![Workforce and Safety!](https://files.oaiusercontent.com/file-Xiw2QHwmdbpJsY3M2yn8JM?se=2024-12-10T19%3A39%3A04Z&sp=r&sv=2024-08-04&sr=b&rscc=max-age%3D604800%2C%20immutable%2C%20private&rscd=attachment%3B%20filename%3D7b4dfcf7-5375-4467-a6df-58c2d1676ed8.webp&sig=rD88H0clo2XVqY1GGvwBaTuMWxi%2B2MeJs4vuM4yq86Q%3D)


# Introduction

The oil and gas industry operates in a complex and high-stakes environment, where workforce dynamics and safety are critical to operational success. This project examines two years of data (2022–2023) from an oil refinery, focusing on employee work hours, overtime, and reported safety incidents. By cleaning and analyzing the data, this study aims to uncover meaningful patterns and trends that shed light on the relationship between workload and safety outcomes. Using visualizations to present the findings, the project provides a clear and concise overview of the data, offering insights to guide future discussions on workforce management and safety strategies.

Time to get started!

```{r}
#| label: "setup"
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA, fig.align = "center")
```

## Loading Necessary Packages

```{r}
#| label: "needed"

# For data importing
library(readxl)
# For data manipulation and tidying
library(tidyverse)
library(stringr)

# For date and time handling
library(lubridate)

```

## Importing Data

The dataset for this project was sourced from an [Excel file](https://docs.google.com/spreadsheets/d/11NnYb1Ubl1xM4rzHecbwei9GDXQHKv-5/edit?gid=660108890#gid=660108890) and imported into R using the read_excel() function from the readxl package. The imported data was then inspected and prepared for analysis through a series of cleaning and transformation steps to ensure accuracy and usability.

```{r}
df <- read_excel("~/git_repos/solanillam/Project_2/Raw_Data/Copy of df_overall_final_V1.xlsx")
```

Ok, let's take a look at the data set.

### Data Structures and Variables